Signatures of Foreground Sky in Power Spectra of Redshifted Neutral Hydrogen from the Epoch of Reionization

### 2014-09-03 foreground_delay_spectrum_modeling.tex

We characterize signatures of foreground radio emission and instrument configuration on the observed power spectra of redshifted H{\sc i} 21~cm line emission from the epoch of reionization (EoR) using an antenna pair based ``delay spectrum'' analysis technique. We use the 128-tile Murchison Widefield Array (MWA) instrument configuration for our studies. % Our analysis is centered around 185~MHz with 30.72~MHz bandwidth at 80~kHz frequency resolution. 
We simulate the delay spectrum response of interferometers, for the first time, to an all-sky foreground model consisting of diffuse and compact Galactic and extragalactic emission. By tying model parameters to those of a subset of first season of observing with the MWA, our simulations agree with the observed data to within expected uncertainty. A wedge--shaped region of foreground contamination and a spillover from this contamination into relatively foreground--free regions ({\it EoR window}) in the delay spectrum due to spectral properties of foreground emission and the instrument is confirmed. Due to shortening of antenna spacings caused by projection effects, diffuse emission produces edge--heavy {\it two--pronged fork}--like features even on wide antenna spacings, hitherto not predicted. Features in the inner regions of the wedge are predominantly due to compact foreground objects. This results in a {\it pitchfork}--shaped delay spectrum. The bright Galactic center severly contaminates the {\it EoR window} on certain antenna spacings in certain observations. For such cases, we present an effective practical tool for screening interferometers with high foreground contamination and mitigating their effects on EoR data analysis.

@@

One of the key challenges in the statistical detection of power spectrum of the redshifted H{\sc i} signal arises from the contamination by Galactic and extragalactic foregrounds \citep[see, e.g.,][]{dim02,zal04,fur06}. \citet{mor04} show that the inherent isotropy and symmetry of the EoR signal in frequency and spatial wavenumber ($k$) space make it distinguishable from sources of contamination which lack such symmetry. Foreground removal techniques rely on the spectral smoothness of foreground emission \citep{mor06,bow09,liu11,par12,dil13}.

Precise understanding of these foreground signatures is necessary in order to remove foreground systematics from the measured power spectrum. Considerable effort has been made in mapping the residual foreground signatures in observed power spectrum \citep{thy13,pob13,mor12,tro12,dat10,bow09} after foreground subtraction.

In this paper, in one of the most detailed characterizations to date, using an interferometer (antenna pair) based delay spectrum technique, we demonstrate the signatures of unsubtracted diffuse and compact foreground models in delay spectrum, a quantity closely related to the sought power spectrum. The simplicity of this technique and usage of basic interferometric visibility data fourier transformed along frequency (delay spectrum) on individual interferometers offer very effective diagnostics for scheduling observations and data analysis, and an insightful view onto foreground signatures in the EoR H{\sc i} power spectrum. We explore the diversity in these signatures as a function of observation and instrument parameters and confirm our findings with data obtained from the MWA. This simple technique allows us to predict foreground signatures, hitherto not predicted.

We have organized this paper as follows. \S\ref{sec:delay-spectrum} defines the delay spectrum and its relation to the power spectrum. \S\ref{sec:instrument} and \S\ref{sec:obsparms} describe the MWA instrumental parameters and observation parameters used in this study respectively. Analysis of data obtained with the MWA is described in \S\ref{sec:data-analysis}. In \S\ref{sec:modeling}, we describe the building blocks of our delay spectrum modeling including the models used for the antenna power pattern and foreground. Here, we also establish that the results of modeling and data analysis agree with each other. In \S\ref{sec:delay-spectrum-analysis}, we present a detailed analysis of the signatures of foregrounds on the modeled delay spectrum and its dependence on instrumental and observation parameters. In \S\ref{sec:fg-grading}, we use our understanding to grade interferometers based on the amount of foreground contamination estimated by the delay spectrum technique. In \S\ref{sec:summary}, we present our summary and conclusions.

>>>>

 Detection and measurement of 21cm HI emission from the Epoch of Reionization, at redshifts above 6, is limited primarily by foreground emission. With much of the information coming from the spectral line that maps to time and line of site distance, foregrounds affecting the spectral domain are deserving of close attention, particularly those that are due to the instrument because they are in theory removable.  Here we investigate in detail the chromatic appearance of flat spectrum foregrounds imprinted by the instrumental transfer function. Using the delay spectrum technique to measure power spectrum contamination on a per-baseline basis we demonstrate that the foreground signatures which have the largest impact on the HI signal arise from power entering far from the center of the field of view.  Comparing data from recent Murchison Widefield Array observations with simulations broken down by sky position and type of emission, we identify smooth emission near the horizon as the largest contributing factor, followed by bright sources in the same locations. The observation that these contaminants only effect a fraction of the baselines leads us to propose that by down-weighting certain baselines at certain times, a large fraction of foreground contamination could be mitigated.

@@

Perhaps the key challenge in the statistical detection of power spectrum of the redshifted H{\sc i} signal arises from the contamination by Galactic and extragalactic foregrounds \citep[see, e.g.,][]{dim02,zal04,fur06}. \citet{mor04} show that the inherent isotropy and symmetry of the EoR signal in frequency and spatial wavenumber ($k$) space make it distinguishable from sources of contamination which are isolated to certain $k$ modes by virtue of their inherent spectral smoothness\citep{mor06,bow09,liu11,par12,dil13}.  Considerable effort has been made towards understanding the $k$-space behavior foreground signatures in the observed power spectrum \citep{thy13,pob13,mor12,tro12,dat10,bow09}.  A general model of how smooth spectrum foregrounds are scattered to higher $k$-modes by the wide-field response of the instrument, into the so-called `wedge',  has emerged that provides a high level explanation for the observed foreground power spectrum. The conservative foreground strategy that has developed alongside this work is to throw out $k$-modes which could be contaminated (Avoidance).  The more aggressive option is to subtract a sky model and regain access to modes lost by Avoidance. In both cases one large unknown is which parts of the sky are most critical to either Avoid or Subtract. Here we will focus primarily on the Avoidance strategy, asking which foreground components are most likely to ``leak'' from foreground modes to Reionization modes.

Foregrounds work their way into the power spectrum by either having intrinsic spectral variability, high instrumental chromaticity, or through approximations in the estimating the power spectrum. Here we are focusing on the instrumental effects. To de-couple from possible analysis effects, such as those pointed out by \citet{Haz13}, we compute power spectra using the per-baseline approach of \citet{pob13,arp14}.  This approximates the power spectrum as the fourier transform of the spectra output by the correlator.

 Matching real observations from the Murchison Widefield Array Epoch of Reionization observing program with instrumental simulations of the entire known sky we examine the signature of the power spectrum of each baseline.  We confirm that the foregrounds most likely to obfuscate the Epoch of Reionization power are not caused by emission central field of view but rather from near the horizon. We also find that the smooth galactic emission plays a larger role then previously thought. We quantify this contamination by breaking the simulations down by sky region and type of emission. Using these we arrive at a new method for minimizing the contribution of bright foregrounds which uses prior knowledge of the sky to downweight badly contaminated baselines.

In \S\ref{sec:delay-spectrum} we introduce the delay spectrum technique, then in  \S\ref{sec:instrument}, \S\ref{sec:obsparms}, and \S\ref{sec:data-analysis} we describe the MWA setup, summarize the observing parameters and present the resulting data.  Simulations of these observing parameters are described in \S\ref{sec:modeling} and analyzed by sky location, and type of emission in \S\ref{sec:delay-spectrum-analysis}.  \S\ref{sec:discussion} offers a discussion along with a basic comparison to other current and future instruments. \S\ref{sec:conclusion} offers conclusions and an initial description of a more precise foreground avoidance technique.
